---
title: "PML Course Project"
author: "Rena Ho"
date: "2025-04-30"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
library(randomForest)
```

## 1. Overview

Using data from accelerometers on various parts of the body, we aim to predict how well participants perform barbell lifts. We use the `classe` variable from the training data as our target and build a classification model using the Random Forest algorithm. The final model is evaluated on a validation set, and predictions are made for a test set.

## 2. Load and Prepare Data

```{r load-data}
training_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testing_url  <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(url(training_url))
testing  <- read.csv(url(testing_url))
```

```{r clean-data}
# Remove columns with mostly NA values and near-zero variance predictors
training <- training[, colSums(is.na(training)) < 0.95 * nrow(training)]
training <- training[, -nearZeroVar(training)]
training <- training[, -(1:7)]  # Remove ID and timestamps
```

## 3. Partition Data

```{r split-data}
set.seed(123)
inTrain <- createDataPartition(training$classe, p = 0.7, list = FALSE)
trainSet <- training[inTrain, ]
valData  <- training[-inTrain, ]
```

## 4. Model Building

```{r model-fit}
control <- trainControl(method = "cv", number = 5)
modelRF <- train(classe ~ ., data = trainSet, method = "rf", trControl = control)
```

## 5. Model Evaluation

```{r evaluation}
predRF_val <- predict(modelRF, newdata = valData)
valData$classe <- factor(valData$classe)
confusionMatrix(predRF_val, valData$classe)
```

## 6. Expected Out-of-Sample Error

```{r error}
accuracy <- postResample(predRF_val, valData$classe)[[1]]
error <- 1 - accuracy
cat("Expected out-of-sample error:", round(error, 4))
```

## 7. Predict on Test Set

```{r final-prediction}
finalPred <- predict(modelRF, newdata = testing)
finalPred
```

## 8. Variable Importance

```{r var-imp}
varImpPlot(modelRF$finalModel)
```

Using Random Forest and 5-fold cross-validation, we achieved strong predictive performance. The model generalizes well, as shown by the validation results and low estimated out-of-sample error. The predictions for the 20 test cases are output above.

## 9. Conclusion

The Random Forest model trained on cleaned accelerometer data achieves high classification accuracy. The expected out-of-sample error is low (~0.006), and model performance is validated. Test set predictions are generated and ready for submission. Variable importance suggests features from the forearm and dumbbell sensors are most influential.